\documentclass{assets/fieldnotes}


\title{Kono (Sierra Leone)}
\author{LING3020/5020}
\date{University of Pennsylvania, Spring 2025\\02/19/2025 Morphology II}

\setcounter{secnumdepth}{4} %enable \paragraph -- for subsubsubsections

\begin{document}

\maketitle
\tableofcontents

\jal{think about simple sentences, topics we've learned are likely naturally spoken of in Kono, coordinate with each other; try pronominal and simple nominal subjects   }

9\section{Nonverbal predication: AdjPs (Wesley)}

\wml{
\begin{itemize}
    \item Seems like emotion-related stage-level adjectives are expressed using body-part idioms, while type-level adjectives are actual adjectival predicates.
    \item Adjectival predicates seem to show morphological subject marking similar to the inalienable possessive series, although I only managed to elicit third-person items with nominal subjects.
    \item Attributive adjectives seem to be marked by -ma, while predicative ones do not bear this marker, only the subject marker.
\end{itemize}


\exg. ̩ḿ̩-bwó dí\\
\textsc{1sg}-body good\\
`I am happy.'

\exg. ̩ḿ̩-bwó fâ énè\\
\textsc{1sg}-body dead ?\\
`I am tired.'

\exg. ɱ̩́-fáà jón déɛ̀\\
\textsc{1sg}-heart poor ?\\
`I am sad.'

\exg. ̩ḿ̩-bwó má-dí\\
\textsc{1sg}-body \textsc{neg}-good\\
`I am not happy.'

\exg. mwókàmá tʃɛ̀ à-bwó má-dí\\
man this \textsc{3sg}-body \textsc{neg}-good\\
`This man is not happy.'

\exg. mwókàmá tʃɛ̀ à-fáà jón déɛ̀\\
man this \textsc{3sg}-heart poor ?\\
`This man is sad.'

\exg. mùsù tʃɛ̀ à-bwó má-dí\\
woman this \textsc{3sg}-body \textsc{neg}-good\\
`This woman is not happy.'

\exg. mwókàmá tʃɛ́-nù à-ɱ-fáà jón-déɛ̀\\
man this-\textsc{pl} \textsc{3-PL}-heart poor-?\\
`These men are sad.'

\exg. mwókàmá tʃẽ̀ĩ à-ɱ-fáà jón-déɛ̀\\
man these \textsc{3-PL}-heart poor-?\\
`These men are sad.'

\exg. mùsú tʃɛ́-nù à-M-bwó má-dì\\
woman this-\textsc{pl} \textsc{3-PL}-body \textsc{neg}-good\\
`These women are not good.'

\exg. mùsú tʃɛ́-nù àɱ-fáà jón déɛ̀\\
woman this-\textsc{pl} \textsc{3pl}-heart poor ?\\
`These women are sad.'

\exg. mwókàmá ɱ̩̀-fáà jón-déɛ̀\\
man \textsc{3pl}-heart poor-?\\
`(Some) men are sad.'

\exg. mùsú ɱ̩̀-fáà jón-déɛ̀\\
woman \textsc{3pl}-heart poor-?\\
`(Some) women are sad.'\\
\wml{Possible to express `women' with a bare noun plus a nasal prefix on the verb, perhaps a reduced form of the 3pl àɱ.}

\exg. mùsù jã̀sàmá tʃɛ̀\\
woman tall this\\
`this tall woman’

\exg. mùsù tʃɛ̀ jã̀sã̀\\
woman this tall\\
`This woman is tall.’

\exg. mùsù tʃɛ́-nù ã̀-jã̀sà\\
woman this-\textscpl{pl} \textsc{3pl}-tall\\
`These women are tall.'

\wml{Distinguishing predicative vs. attributive AdjPs}

\exg. mùsù jã̀sàmá tʃɛ̀ á-ɲí\\
woman tall this \textsc{3sg}-beautiful\\
`This tall woman is beautiful.'

\exg. mùsù ɲímá tʃɛ̀ á-jã̀sã̀\\
woman beautiful this \textsc{3sg}-tall\\
`This beautiful woman is tall.'

\wml{Rising intonation at the end of the DP---is that a hanging topic?}



\section{Nonverbal predication: PPs (Mingyang)} 
\ex. tʃénè\\
    `house'

\exg. ḿ-bé tʃén-à\\
    1SG-NPST house-P\\
    I am in the house.

\mb{It might be the case that the postposition for `in' -a comes after tʃénè, so the result is tʃénà.}
   
\exg. é tʃén-à\\
    2SG.NPST house-P\\
    You(sg.) are in the house.

\exg. w-é tʃén-à\\
    2PL-NPST house-P\\
    You(pl.) are in the house.

\ex. ɛ̀ tʃén-à\\
    3SG.NPST house-P\\
    She/he is in the house.

\exg. à-m-bè tʃén-à\\
    3-PL-NPST house-P\\
    They are in the house.

\exg. bèé tʃénà\\
    1.PL.EXCL house\\
    We(excl., du./pl.) are in the house.

\exg. mw-ɛ̃̀  tʃénà\\
    1INCL.DU-NPST house\\
    We(incl., du.) are in the house.

\exg. mw-ɛ̃̂  tʃénà\\
    1INCL-NPST house\\
    We(incl., pl.) are in the house.

\ex. jàí tʃénà\\
    Jai house\\
    Jai is in the house.


\ex. kàâ\\
    `snake'

\ex. sènê\\
    `stone'

\exg. kàâ ɛ̀ sènê mà.\\
    snake 3SG.NPST stone on\\
    The snake is on the stone.

\mb{I suspect that there is a morpheme è for subjects (not sure if it's only for animate subjects), and kàâ è sounds like kàê.}

\exg. kàâ ɛ̀ sènê kɔ́ɔ́.\\
    snake 3SG.NPST stone under\\
    The snake is under the stone.

\exg. sènê kàâ kɔ́ɔ́.\\
    stone snake under\\
    The stone is under the snake.

\exg. kàâ ɛ̀ sènê jɔ̂.\\
    snake 3SG.NPST stone in.front.of\\
    The snake is in front of the stone.

\exg. kàâ ɛ̀ sènê bà.\\
    snake 3SG.NPST stone behind\\
    The snake is behind the stone.

Pronouns for `in the house':
\begin{center}
    \begin{tabular}{|c|c|c|c|}
    \hline
    Person/Number & SG & DU & PL \\ \hline
    1 & mbé & mùè(incl.) & mùê(incl.)/bèé(excl.) \\ \hline
    2 & éé & N/A & wéé \\ \hline
    3 & ɛ̀ & N/A & àmbè \\ \hline
    \end{tabular}
\end{center}
Prepositions:
\begin{center}
    \begin{tabular}{|c|c|}
    \hline
    on & mà\\ \hline
    under & kɔ́ɔ́\\ \hline
    in front of & jɔ̂\\ \hline
    behind & bà\\ \hline
    \end{tabular}
\end{center}

\section{Nonverbal predication: NPs (Daniel)}
\ex. kámw\textipa{\`E}\\
teacher

\exg. kámw\textipa{\`E} mu n-á\\
teacher COP 1SG-POSS\\
`I am a teacher'\\
\ds{Note order, pred-first -- is the morpheme at the end agreement or pron?}

\exg. kámw\textipa{\`E} mu j-á\\
teacher COP 2SG-POSS\\
`You (sg.) are a teacher'

\exg. kámw\textipa{\`E} mu w-á\\
teacher COP 2PL-POSS\\
`You (pl.) are teachers'

\exg. kámw\textipa{\`E} mu ná-à\\
teacher COP 1EXCL-POSS\\
`We (dual excl.) are teachers'

\exg. kámw\textipa{\`E} mu mwa\\
teacher COP 1PL.EXCL\\
`We (pl excl.) are teachers


\exg. kà-mó\\
teach-person\\
`teacher’\\

\exg. Kai mu kámw\textipa{\`E\`E} na\\
Kai COP teacher-? 3SG?\\
`Kai is a teacher'
\ds{*Kai kámw\textipa{\`E\`E} na mu -- copula cannot be postposed}\\
\ds{Lengthening of predicate?}

\exg. Kámw\textipa{\`E\`E} mu Kai-à\\
teacher-? COP Kai-3SG.POSS\\
`Kai is a teacher' (unclear if this is equivalent)
\ds{*Kai kamw\textipa{\`E\`E}(na) -- copula cannot be dropped}


\exg. Kai mu kámw\textipa{\`E} \textipa{\textltailn}imá na\\
Kai COP teacher good 3SG?\\
`Kai is a good teacher'

\exg. táá\\
calabash\\

\exg. táá mu dawn-fen-àà na\\
calabash COP food-thing-? 3SG?\\
`A calabash is food'

\exg. táá mu dawn-f\textipa{\~E} \textipa{tS\~edE} na\\
calabash COP food-thing good 3SG?\\
`A calabash is good food'


\exg. táá mu dawn-f\textipa{\~E} \textipa{\textltailn}imá ná\\
calabash COP food-thing good 3SG?\\
`A calabash is good food'

\exg.  táá \textipa{tSe-nu} mu dawn-f\textipa{\~E} \textipa{tS\~edE} ana\\
calabashes this-PL COP food-thing good 3PL\\
`These calabashes are good food' (?)

\exg. táá-nu mu dawn-f\textipa{\~E} \textipa{tS\~edE} ana\\
calabashes COP food-thing good 3PL\\
`Calabashes are good food' (?)



\section{Intransitive verbal predicates: Past (Chun-Hung)}

\jal{use an adverb (e.g. yesterday) to make sure you're getting the temporal interpretation you want}

\chs{unergatives (others: run/walk, swim, sneeze, sing ...)}

\exg. K\`{a}\`{i} \`{a} d\`{i}-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
Kai 3SG.PST cry-? yesterday \\
`Kai cried yesterday.' \chs{di with a slightly rising tone?}
 
\exg. D\'{e}n\'{e}-t\textipa{S}\`{\textipa{E}} \`{a} d\`{i}-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
child-DEM 3SG.PST cry-? yesterday \\
`This child cried yesterday.'

\exg. N-\'{a} d\`{i}-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
1SG.PST cry-? yesterday \\
`I cried yesterday.'

\exg. J-\'{a} d\`{i}-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
2SG-PST cry-? yesterday \\
`You (sg.) cried yesterday.' 

\exg. \`{A} d\`{i}-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
3SG.PST cry-? yesterday \\
`He/She cried yesterday.'

\exg. N\`{a}-\'{a} d\`{i}\'{e}n-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
1EXCL-PST cry-? yesterday \\
`We (exclusive, dual/plural) cried yesterday.'

\exg. M\`{ɔ}\'{a} d\`{i}\'{e}n-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
1INCL.DU-PST cry-? yesterday \\
`We (inclusive, dual) cried yesterday.'

\exg. M\'{ɔ}\'{a} d\`{i}\'{e}n-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
1INCL-PST cry-? yesterday \\
`We (inclusive, plural) cried yesterday.'

\exg. W\'{a} d\`{i}\'{e}n-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
2PL-PST cry-? yesterday \\
`You (pl.) cried yesterday.' 

\exg. \`{A}-n-\'{a} d\`{i}\'{e}n-t\textipa{S}\`{\textipa{E}} k\`{u}n\`{u}. \\
3-PL-PST cry-? yesterday \\
`They cried yesterday.'

\chs{unaccusative (others: come, leave, die ...)}
\jal{arrive/come/leave depend on the language having these as directed motion verbs, not all languages do. Since we don't have much time and die is a bit morbid, perhaps try fall?}

\exg. K\`{a}\`{i} b\'{e}\`{a} k\`{u}n\`{u}. \\
Kai fall yesterday \\
`Kai fell yesterday.'

\exg. D\'{e}n\'{i}n-t\textipa{S}\`{\textipa{E}} b\'{e}\`{a} k\`{u}n\`{u}. \\
child-DEM fall yesterday \\
`This child fell yesterday.'

\exg. \'{M} b\'{e}\`{a} k\`{u}n\`{u}. \\
1SG fall yesterday \\
`I fell yesterday.' \chs{*N\'{a} for 1SG cannot be used here.}

\exg. \'{I} b\'{e}\`{a} k\`{u}n\`{u}. \\
2SG fall yesterday \\
`You (sg.) fell yesterday.' \chs{*J\'{a} for 2SG cannot be used here.}

\exg. \`{A} b\`{e}\'{a} k\`{u}n\`{u}. \\
3SG fall yesterday \\
`He/She fell yesterday.'

\exg. \`{M} b\`{e}\'{a} k\`{u}n\`{u}. \\
1EXCL fall yesterday \\
`We (exclusive, dual/plural) fell yesterday.'

\exg. M\`{o} b\'{e}\`{a} k\`{u}n\`{u}. \\
1INCL.DU fall yesterday \\
`We (inclusive, dual) fell yesterday.'

\exg. M\'{o} b\'{e}\`{a} k\`{u}n\`{u}. \\
1INCL fall yesterday \\
`We (inclusive, plural) fell yesterday.'

\exg. W\'{o} b\'{e}\`{a} k\`{u}n\`{u}. \\
2PL fall yesterday \\
`You (pl.) fell yesterday.'

\exg. Ã́ b\'{e}\`{a} k\`{u}n\`{u}. \\
3PL fall yesterday \\
`They fell yesterday.' 

\section{Transitive verbal predicates: Past (Joey)} 

\ex. dàó̃\\
    `eat'\\

\ex. dàùnɛ̂\\
    `eat'\\


\exg. \textipa{sw\'e\`e} \\
`meat'\\
  

\jf{When taken with data from the previous section, there appears to be a split-S ergative/absolutive system}\\

\exg. n-á swéè dàó̃ kùnù.\\
1SG.PST meat eat yesterday \\%
    `I ate meat yesterday'

\exg. j-á swéè dàó̃ kùnù\\
2SG-PST meat eat yesterday \\%
    `You (sg.) ate meat yesterday'

\exg. á swéè dàó̃ kùnù\\
3SG.PST meat eat yesterday \\%
    `He/She ate meat yesterday'

\exg. kàì á swéè dàó̃ kùnù\\
Kai AUX.PST meat eat yesterday \\%
    `Kai ate meat yesterday'

\exg. nà-á swéè dàó̃ kùnù\\
1EXCL-PST meat eat yesterday \\
    `We (exclusive, pl.) ate meat yesterday'

\ex. Same as exclusive, plural version\\
    `We (exclusive, dual) ate meat yesterday'

\exg. múá swéè dàó̃ kùnù\\
1DU.INCL meat eat yesterday \\
    `We (inclusive, dual) ate meat yesterday'

\exg. múà swéè dàó̃ kùnù\\
1PL.INCL meat eat yesterday \\
    `We (inclusive, pl.) ate meat yesterday'

\exg. w-á swéè dàó̃ kùnù\\
2SG-PST meat eat yesterday \\
    `You (pl.) ate meat yesterday'

\exg. à-n-á swéè dàó̃ kùnù\\
3-PL-PST meat eat yesterday \\
    `They ate meat yesterday'

\exg. sàa ɱfea kàìjâ à-n-á swéè dàó̃ kùnù\\
Saa and Kai.? 3-PL-PST meat eat yesterday \\
    `Saa and Kai ate meat yesterday'

\ex. íễː\\
    `see'

\exg. n-á í íễː kùnù\\
1SG.PST 2SG see yesterday \\
    `I saw you (sg.) yesterday'

\exg. n-á à íễː kùnù\\
1SG.PST 3SG see yesterday \\
    `I saw him yesterday'

\exg. n-á sàà à íễː kùnù\\
1SG.PST Saa 3SG see yesterday \\
    `I saw Saa yesterday'

\exg. n-á wó íễː kùnù \jf{(When said quickly, can sound like ná wé íễː kùnù)}\\
1SG.PST 2PL see yesterday \\
    `I saw you (pl) yesterday'

\exg. n-á á̃ íễː kùnù \jf{(When said quickly, can sound like ná  ánd͡ʒ íễː kùnù)}\\
1SG.PST 3PL see yesterday \\
    `I saw them yesterday'

\exg. n-á sàà nì kàì̃ íễː kùnù\\
1SG.PST Saa and Kai.3PL see yesterday \\
    `I saw Saa and Kai yesterday'

\exg. n-á sàà ɱfea kàì̃ íễː kùnù\\
1SG.PST Saa and Kai.3PL see yesterday \\
    `I saw Saa and Kai yesterday'



\section{Nonverbal predication: Past (Lex)}%Lex


\exg. ʃopi\\
store\\
`store'

\exg. ʃopjɔ̀, \jf{ Sometimes Anthony pronounced it sopjɔ̀}\\
at the store\\
`at the store'

\exg. kùnù\\
yesterday\\
`yesterday'


\exg. m-fààyondɛ̀ níkùnù, m-bwɔ̀nímádìkùnù\\
1SG was sad yesterday\\
`I was sad, yesterday.'

\exg. m-fààyondɛ̀ níkùnù ʃopjɔ̀ \\
1SG was sad yesterday in the store\\
I was sad in the store, yesterday.\chs{( ʃopjɔ̀kùnù, m-fààyondɛ̀ní ) means the same thing}

\exg. m-fààyondɛ̀ níkùnùtʃɛ̀ná \\
1SG. was sad yesterday in the house\\
I was sad in the house, yesterday.\chs{( tʃɛ̀nákùnù, m-fàà yondɛ̀ní ) means the same thing}

\exg. í-fààyondɛ̀ níkùnù\\
2SG. were sad yesterday\\
`You (sg.) were sad, yesterday.'

\exg. á-fààyondɛ̀ níkùnù\\
3SG. was sad yesterday\\
 ` He was sad, yesterday.'
 
\exg. á-fààyondɛ̀ níkùnù\\
3SG. was sad yesterday\\
`She was sad, yesterday'

\exg. áŋ-fààyondɛ̀ níkùnù\\
3PL. were sad yesterday\\
`They were sad, yesterday'

\exg. mm-fààyondɛ̀ níkùnù\\ 
1DU.EXCL. were sad yesterday\\
`We (\textsc{du.excl}) were sad, yesterday' %\jal{Not sure how to transcribe the difference in m pronunciation (sounded like some kind or rising pitch and lengthened m) of "I was sad" vs "We were sad"

\exg. mɔ-fààyondɛ̀ níkùnù\\
1DU.INCL. were sad yesterday\\
 `We (\textsc{du.incl}) were sad, yesterday' 

\ex. `We (\textsc{Pl.excl}) were sad, yesterday'

\exg. mɔ́-fààyondɛ̀ níkùnù\\
1INCL. were sad yesterday\\
 `We (\textsc{Pl.incl}) were sad, yesterday'


%\exg. bànù\\
last year\\
`last year'

%\exg. kà-mó\\
teach-person\\
`teacher’\\


\section{Nonverbal predication: Future (Jan)} 

\jal{it's useful to have minimal pairs with other tenses; build on what we have. -- so use what other people have gotten before you  Use adverbs like "tomorrow" etc to help the temporal interpretation.}

\ex. kúíì \\
`leopard' 


\exg. kúíì cɛ̀ \\
leopard this \\
`this leopard'

\ex. sáándʒé \\
`soon' (immediate)

\ex. flòflò \\
`soon' (immediate)

\exg. kwíì tʃè koáà wà \\
leopard this big? FUT \\
`This leopard will be big.'

\exg. kwì tʃè kʷóáá wà sáándʒé \\
leopard this big? FUT soon \\
`This leopard will be big soon.'

\exg. kwì tʃɛ̀ ɛ́ dó éàwà \\
leopard this 3SG? small FUT? \\
`This leopard will be small.'

\exg. sáán tɛ̀mà pwì tʃà \\
Sahr POSS.FUT leopard this \\
`This leopard will be Sahr's.'

\exg. sàán tàmù \\
Sahr POSS-COP \\
`This is Sahr's' (certainty?)

\ex. sínà \\
`tomorrow' 

\exg. kwi tʃè-è màwàŋ kònè kóò sìnà \\
leopard this-? LOC.FUT tree under tomorrow \\
`This leopard will be under the tree tomorrow.' %\jal{doesn't strike me as very natural? think about a situation where you'd want to say something}

\ex. ii \\
`river'

\exg. kwí tʃè-è màwà íì fè sìnà  \\
leopard this-? LOC.FUT river beside tomorrow \\
`This leopard will be by the river tomorrow.'

\exg. kwí tʃè màwà nè sìnà \\
leopard this LOC.FUT here tomorrow \\ 
`This leopard will be here tomorrow.'


\exg. kwí tʃè dàònfènà són dá wà sìnà \\
leopard this food POSS? POSS FUT tomorrow \\
`This leopard will have food tomorrow.'


\ex. m-bátʃɛ̀nɛ́ \\ 
`neighbor'

\exg. káé màwà mbátʃènà \\
Kai POSS-FUT neighbor \\
`Kai will be my neighbor.'

\ex. gbándìmá \\
`soon' (not immediate)

\exg. káé màwà mbátʃènà àgbándìmá\\
Kai POSS-FUT neighbor soon \\
`Kai will be my neighbor soon.'

\jmt{Here, Anthony noted that using [sáándʒé] would be incorrect, since the `immediate' meaning does not fit.}

\exg. káé màwà sáà batʃènà \\
Kai POSS.FUT Sahr neighbor \\
`Kai will be Sahr's neighbor.'

\exg. káé bàtʃènè mà són dà-wà \\
Kai neighbor POSS POSS POSS-FUT \\
`Kai will have a neighbor.'


\section{Verbal predication: Progressive (Alex)} %Alex

\exg.
ḿ-bé        béjà \\
1SG.NPST   fall \\%
`I am falling.'

\exg.
é         béjà \\
2SG.NPST   fall \\%
`You(sg) are falling.'

\exg.
ɛ̀         bèjà \\
3SG.NPST   fall \\%
`He/she is falling.'

\exg.
mw-ɛ̃̀    bèjà \\
1INCL.DU-NPST   fall \\%
`We(dual) are falling.'

\exg.
mw-ɛ̃̂         bèjà \\
1INCL-NPST  fall \\%
`We(incl) are falling.'

\exg.
m-bè         bèjà \\
1EXCL-NPST   fall \\%
`We(excl) are falling.' \label{We(excl) are falling}

\alex{not sure about the tone on \ref{We(excl) are falling}.}

\exg.
w-é        béjà \\
2PL-NPST   fall \\%
`You(pl) are falling.'

\exg.
à-m-bè      bèjà \\
3-PL-NPST   fall \\%
`They are falling.'

\exg.
n-a    jaɡbasi   a     beja \\
1SG.PST   onion     OBJ   fall \\%
`I dropped the onion.'

\exg.
ḿ-bé      jaɡbasi   a     beja \\
1SG.NPST   onion     OBJ   fall \\%
`I am dropping the onion.'

\exg.
ḿ-bé      jaɡbasi   a     beja   waN   timotima   \\
1SG.NPST   onion     OBJ   fall   FOC   every.time \\%
`I always drop the onion.' \label{I always drop the onion}

\exg.
timotima     ḿ-bé       jaɡbasi   a     beja   waN \\
every.time   1SG.NPST   onion     OBJ   fall   FOC \\%
`I always drop the onion.'

\alex{The non-past/progressive/imperfective form is also used to express habituals, as shown by \ref{I always drop the onion}.}

\alex{Nice unaccusative/transitive (inchoative/causative) minimal pair with \textit{beja} `fall'.}

\alex{Object marker /a/ shows up in both past and non-past clauses}

\exg.
é         jaɡbasi   a     beja \\
2SG.NPST   onion     OBJ   fall \\%
`You(sg) are dropping the onion.'

\exg.
ɛ̀         jaɡbasi   a     beja \\
3SG.NPST   onion     OBJ   fall \\%
`He/she is dropping the onion.'

\exg.
a     jaɡbasi   tʃɛ-n-a       beja \\
3SG.PST   onion     this-PL-OBJ   fall \\%
`He/she dropped these onions.'

\exg.
ɛ̀         jaɡbasi   tʃɛ-n-a       beja \\
3SG.NPST   onion     this-PL-OBJ   fall \\%
`He/she is dropping these onions.'


\section{Verbal predication: Future (Giang)} %Giang


\g{Unaccusatives, most likely:}

\ex. báj.á.mà\\
``Baiama''

\exg. Sàà é ná wá Bájámà sénà.\\
Saa ?(3SG.NPST??) come \textsc{fut} Baiama tomorrow\\
Saa will arrive in Baiama tomorrow. 
\g{is is e, does it carry tone?}


\exg. m-bé nà wà Bájámà sénà.\\
1\textsc{SG-NPST} come \textsc{fut} Baiama tomorrow\\
I will arrive in Baiama tomorrow.

\exg. éé nà wà Bájámà sénà.\\
You (2SG.NPST??) come \textsc{fut} Baiama tomorrow\\
You will arrive in Baiama tomorrow.

\exg. wéé nà wà Bájámà sénà.\\
2\textsc{PL.NPST} come \textsc{fut} Baiama tomorrow\\
You(pl) will arrive in Baiama tomorrow.

\exg. mw-è ná wà Bájámà sénà.\\
1.\textsc{INCL.DU-NPST}.\textsc{incl} come \textsc{fut} Baiama tomorrow\\
We(dual, inclusive) will arrive in Baiama tomorrow.

\exg. m-bé ná wà Bájámà sénà.\\
1.\textsc{EXCL-NPST}.\textsc{exl} come \textsc{fut} Baiama tomorrow\\
We(dual, exclusive) will arrive in Baiama tomorrow.

\exg. á-m-bè ná wà Bájámà sénà.\\
3-\textsc{PL-NPST} come \textsc{fut} Baiama tomorrow\\
They will arrive in Baiama tomorrow.


\g{Unergative}
\ex. tómbwè\\
to dance

\ex. íí-kuma\\
by the river

\ex. sànàmà bwódjè\\
New years?

\exg. Sàà é tómbwè dòndá-wà ííkùmà sénà\\
Saa (3SG.NPST)? dance ? by.the.river tomorrow\\
Saa will dance by the river tomorrow. 

\exg. m-bé tómbwè dòndá-wà ííkùmà sénà\\
1\textsc{SG-NPST} ? dance ? by.the.river tomorrow\\
I will dance by the river tomorrow.\\
\g{Is dòndáwà another way to mark \textsc{fut}?}

\exg. mw-è tómbwé dòndá-wa ííkùmà sénà\\
1\textsc{INCL.DU-NPST} ? dance ? by.the.river tomorrow\\
We(dual, incl) will dance by the river tomorrow.


\g{Transitive}
\ex. fí.án.né\\
broom, sweep\\
\g{né is nominalization.}

\exg. Sàà é cénà fíándà sénà\\
Saa ?(3SG.NPST) house sweep tomorrow\\
Saa will sweep the house tomorrow.\\
\g{house might be house+obj marker.}


\exg. é cénà fíándà sénà\\
2\textit{SG.NPST} house sweep tomorrow\\
You will sweep the house tomorrow.

\exg. é cénà fíándà wá sénà\\
2\textit{SG.NPST} house sweep have.to tomorrow\\
You(sg) will have to sweep the house tomorrow.

\exg. W-é cénà fíándà sénà\\
2\textit{PL-NPST} house sweep tomorrow\\
You(pl) will sweep the house tomorrow.

\exg. mw-è cénà fíándà sénà\\
1.\textit{INCL.DU-NPST} house sweep tomorrow\\
We(dual, incl) will sweep the house tomorrow.

\exg. m-bé cénà fíándà sénà\\
1.\textit{EXCL-NPST} house sweep tomorrow\\
We(dual, incl) will sweep the house tomorrow.

\exg. mw-ê cénà fíándà sénà\\
1.\textit{INCL-NPST} house sweep tomorrow\\
We(pl, incl) will sweep the house tomorrow.

\exg. á-m-bè cénà fíándà sénà\\
3-\textsc{PL.NPST} house sweep tomorrow\\
They will sweep the house tomorrow.

\end{document}